Monte Carlo Expectation Maximization for RNAseq

### 2015-06-16 slides/EM/em.Rnw
through mcmcem

\title{Monte Carlo Expectation Maximization for RNAseq}
\author[Jarad Niemi]{Dr. Jarad Niemi}
\institute[Iowa State]{Iowa State University}
\date{\today}

@@

<<chunk_options, echo=FALSE, message=FALSE>>=
opts_chunk$set(fig.width=6, fig.height=5, out.width='.8\\linewidth', fig.align='center', size='tiny')
library(plyr)
library(ggplot2)
library(reshape2)
@

@@

<<normal_hierarchical_mean, cache=TRUE, echo=FALSE>>=
G = 1000
truth = data.frame(mu = 0, tau = 1)
d = ddply(data.frame(gene=1:G, theta = with(truth, rnorm(G, mu, sqrt(tau)))),
          .(gene), function(x) data.frame(gene=x$gene, y=rnorm(4, x$theta)))
sm = ddply(d, .(gene), summarize, n = length(y), ybar = mean(y))
mu = tau = 1 # Initial values
n_iter = 10
keep = data.frame(iteration = 1:n_iter, mu = rep(NA, n_iter), tau = rep(NA, n_iter))
for (i in 1:n_iter) {
  sm$v = 1/(1/tau+sm$n)
  sm$m = sm$v * (mu/tau + sm$ybar*sm$n)
  mu = mean(sm$m)
  tau = mean(sm$v + (sm$m-mu)^2)
  keep$mu[i] = mu; keep$tau[i] = tau
}
ggplot(melt(keep, id.var="iteration", variable.name="parameter"), 
       aes(iteration, value)) +
  geom_point() + 
  facet_wrap(~parameter, scales="free_y") +
  geom_hline(data=melt(truth, variable.name='parameter'), aes(yintercept=value), color='red')
@

@@

This Monte Carlo approximation to the desired expectation

@@

\[ \begin{array}{rl}
\mu^{(t+1)} &= \frac{1}{MG} \sum_{m=1}^M \sum_{g=1}^G \theta_g^{(m)} \\
\tau^{(t+1)} &= \frac{1}{MG} \sum_{m=1}^M \sum_{g=1}^G \left(\theta_g^{(m)} - \mu^{(t+1)} \right)^2
\end{array} \]
\end{frame}

>>>>

\title[MCMCEM for RNAseq]{Monte Carlo Expectation Maximization for RNAseq}
\author[Jarad Niemi]{Dr. Jarad Niemi}
\institute[Iowa State]{Iowa State University}
\date{\today}

@@

<<chunk_options, echo=FALSE, message=FALSE>>=
opts_chunk$set(fig.width=6, fig.height=5, out.width='.8\\linewidth', fig.align='center', size='tiny')
library(plyr)
library(ggplot2)
library(reshape2)
library(rstan)
@

@@

<<normal_hierarchical_mean_em, cache=TRUE, echo=FALSE>>=
G = 100
truth = data.frame(mu = 0, tau = 1)
d = ddply(data.frame(gene=1:G, theta = with(truth, rnorm(G, mu, sqrt(tau)))),
          .(gene), function(x) data.frame(gene=x$gene, y=rnorm(4, x$theta)))
sm = ddply(d, .(gene), summarize, n = length(y), ybar = mean(y))
mu = tau = 1 # Initial values
n_iter = 10
keep = data.frame(iteration = 1:n_iter, mu = rep(NA, n_iter), tau = rep(NA, n_iter), method="em")
for (i in 1:n_iter) {
  sm$v = 1/(1/tau+sm$n)
  sm$m = sm$v * (mu/tau + sm$ybar*sm$n)
  mu = mean(sm$m)
  tau = mean(sm$v + (sm$m-mu)^2)
  keep$mu[i] = mu; keep$tau[i] = tau
}
@

<<, dependson='normal_hierarchical_mean_em', echo=FALSE>>=
ggplot(melt(keep, id.var=c("iteration","method"), variable.name="parameter"), 
       aes(iteration, value)) +
  geom_point() + 
  facet_wrap(~parameter, scales="free_y") +
  geom_hline(data=melt(truth, variable.name='parameter'), aes(yintercept=value), color='red')
@

@@

The Monte Carlo approximation to the desired expectation is

@@

\[ \begin{array}{rl}
\mu^{(t+1)} &= \frac{1}{G} \sum_{g=1}^G \frac{1}{M}\sum_{m=1}^M \theta_g^{(m)} \\
\tau^{(t+1)} &= \frac{1}{G} \sum_{g=1}^G  \frac{1}{M}\sum_{m=1}^M\left(\theta_g^{(m)} - \mu^{(t+1)} \right)^2
\end{array} \]
\pause 
where $\theta_g^{(m)} \stackrel{ind}{\sim} p\left(\theta_g\left| y_g,\pi^{(t)}\right.\right)$.
\end{frame}

\begin{frame}[fragile]
<<normal_hierarchical_mean_mcem, dependson="normal_hierarchical_mean_em", cache=TRUE, echo=FALSE>>=
M = 1000
mu = tau = 1 # Initial values
keep2 = data.frame(iteration = 1:n_iter, mu = rep(NA, n_iter), tau = rep(NA, n_iter), method='mcem')
for (i in 1:n_iter) {
  sm$v = 1/(1/tau+sm$n)
  sm$m = sm$v * (mu/tau + sm$ybar*sm$n)
  draws = ddply(sm, .(gene), function(x) {
    data.frame(theta = rnorm(M, x$m, sqrt(x$v)))
  })
  mu = mean(draws$theta)
  tau = mean((draws$theta-mu)^2)
  keep2$mu[i] = mu; keep2$tau[i] = tau
}
@

<<dependson=c('normal_hierarchical_mean_em','normal_hierarchical_mean_mcem'), echo=FALSE>>=
ggplot(melt(rbind(keep,keep2), id.var=c("iteration","method"), variable.name="parameter"), 
       aes(iteration, value, color=method)) +
  geom_point() + 
  facet_wrap(~parameter, scales="free_y") +
  geom_hline(data=melt(truth, variable.name='parameter'), aes(yintercept=value))
@
\end{frame}

\subsection{MCMC EM}
\begin{frame}
\frametitle{Markov chain Monte Carlo EM}

It will not always be easy to sample from $p\left(\theta\left|y,\pi^{(t)}\right.\right)$, so we may want to use an MCMC approach to sample from this distribution. \pause The resulting MCMCEM algorithm is identical to the MCEM algorithm, i.e.

\[ \begin{array}{rl}
\mu^{(t+1)} &= \frac{1}{G} \sum_{g=1}^G \frac{1}{M}\sum_{m=1}^M \theta_g^{(m)} \\
\tau^{(t+1)} &= \frac{1}{G} \sum_{g=1}^G  \frac{1}{M}\sum_{m=1}^M\left(\theta_g^{(m)} - \mu^{(t+1)} \right)^2
\end{array} \]
\pause where 
\[ \theta_g^{(m)} \sim p\left(\theta_g\left|y_g,\pi^{(t)}\right.\right) \]
but now $\theta_g^{(m)}$ and $\theta_g^{(m+1)}$ are not independent \pause (they are still independent across $g$).

\begin{frame}[fragile]
<<normal_hierarchical_mean_mcmcem, dependson="normal_hierarchical_mean_em", cache=TRUE, echo=FALSE, results='hide'>>=
normal_hierarchical_mean_model = "
data {
  int<lower=1> n;
  real y[n];
  real mu;
  real<lower=0> s;
  real<lower=0> tau;
}
parameters {
  real theta;
}
model {
  theta ~ normal(mu,sqrt(tau));
  y ~ normal(theta, s);
}
"

m = stan_model(model_code = normal_hierarchical_mean_model)

mu = tau = 1 # Initial values
keep3 = data.frame(iteration = 1:n_iter, mu = rep(NA, n_iter), tau = rep(NA, n_iter), method='mcmcem')
for (i in 1:n_iter) {
  # Run MCMC using rstan
  draws = ddply(d, .(gene), function(x) {
    s = sampling(m, data = list(n = length(x$y), y=x$y, mu=mu, s=1, tau=tau), iter=M/2)
    data.frame(theta = as.numeric(extract(s, 'theta')$theta))
  })

  mu = mean(draws$theta)
  tau = mean((draws$theta-mu)^2)
  keep3$mu[i] = mu; keep3$tau[i] = tau
}
@

<<dependson=c('normal_hierarchical_mean_em','normal_hierarchical_mean_mcem'), echo=FALSE>>=
ggplot(melt(rbind(keep,keep2,keep3), id.var=c("iteration","method"), variable.name="parameter"), 
       aes(iteration, value, color=method)) +
  geom_point() + 
  facet_wrap(~parameter, scales="free_y") +
  geom_hline(data=melt(truth, variable.name='parameter'), aes(yintercept=value))
@

\section{Applications}
\begin{frame}
\frametitle{Negative binomial example}

Let's suppose $y_{gi} \stackrel{ind}{\sim} NB(\theta_g,\psi_g)$, i.e. 
\[ E[y_{gi}] = \eta_g = e^{\theta_g} \quad\mbox{and}\quad V[y_{gi}] = \eta_g + e^{\psi_g}\eta_g \]
\pause 
and we assume a hierarchical distribution for $\theta_g$ and $\psi_g$:
\[ 
\theta_g \stackrel{ind}{\sim} N(\mu_\theta,\tau_\theta) \quad\mbox{and, independently,}\quad \psi_g \stackrel{ind}{\sim} N(\mu_\psi, \tau_\psi).
\]
\end{frame}

\begin{frame}
\frametitle{MCMCEM for negative binomial hierarchical model}

Let $\pi=(\mu_\theta, \tau_\theta, \mu_\psi, \tau_\psi)$, then 
\[ 
E_{\theta,\psi|y,\pi^{(t)}}[\log p(\theta,\psi|\pi)] = E_{\theta|y,\pi^{(t)}}[\log p(\theta|\mu_\theta,\tau_\theta)] + E_{\psi|y,\pi^{(t)}}[\log p(\psi|\mu_\psi,\tau_\psi)]
 \]
 which is maximized with
\[ \begin{array}{rl}
\mu_\theta^{(t+1)} &= \frac{1}{G} \sum_{g=1}^G \frac{1}{M}\sum_{m=1}^M \theta_g^{(m)} \\
\tau_\theta^{(t+1)} &= \frac{1}{G} \sum_{g=1}^G  \frac{1}{M}\sum_{m=1}^M\left(\theta_g^{(m)} - \mu_\theta^{(t+1)} \right)^2 \\
\mu_\psi^{(t+1)} &= \frac{1}{G} \sum_{g=1}^G \frac{1}{M}\sum_{m=1}^M \psi_g^{(m)} \\
\tau_\psi^{(t+1)} &= \frac{1}{G} \sum_{g=1}^G  \frac{1}{M}\sum_{m=1}^M\left(\psi_g^{(m)} - \mu_\psi^{(t+1)} \right)^2
\end{array} \]
\pause where 
\[ \theta_g^{(m)},\psi_g^{(m)} \sim p\left(\theta_g,\psi_g\left|y_g,\pi^{(t)}\right.\right) \]
via MCMC. 
\end{frame}

<<negative_binomial_data, cache=TRUE>>=
G = 100
n = 4
truth = data.frame(mu_theta = 3, tau_theta = 1, mu_psi = 0, tau_psi = 1)
par = data.frame(gene=1:G, 
                 theta = with(truth, rnorm(G, mu_theta, sqrt(tau_theta))),
                 psi   = with(truth, rnorm(G, mu_psi, sqrt(tau_psi))))
d = ddply(par, .(gene), function(x) {
  data.frame(y = rnbinom(n, mu = exp(x$theta), size=1/exp(x$psi)))
})
@

<<negative_binomial_stan, cache=TRUE>>=
negative_binomial_model = "
data {
  int<lower=1> n;
  int<lower=0> count;
  real mu_theta;
  real<lower=0> tau_theta;
  real mu_psi;
  real<lower=0> tau_psi;
}
parameters {
  real theta;
  real psi;
}
model {
  theta ~ normal(mu_theta, sqrt(tau_theta));
  psi   ~ normal(mu_psi  , sqrt(tau_psi  ));
  count ~ neg_binomial_2_log(theta, 1/psi);
}
"
nb_m = stan_model(model_code = negative_binomial_model)
@

<<negative_binomial_mcmcem, dependson=c('negative_binomial_data','negative_binomial_stan'), cache=TRUE, echo=FALSE, results='hide'>>=
M = 1000
mu_theta = 0; tau_theta=1, mu_psi = 0; mu_psi = 1 # Initial values
n_iter = 10
keep = data.frame(iteration = 1:n_iter, 
                  mu_theta = rep(NA, n_iter), 
                  tau_theta = rep(NA, n_iter), 
                  mu_psi = rep(NA,n_iter),
                  tau_psi = rep(NA,n_iter))
for (i in 1:n_iter) {
  # Run MCMC using rstan
  draws = ddply(d, .(gene), function(x) {
    s = sampling(nb_m, data = list(n = length(x$y), y=x$y, 
                                   mu_theta=mu_theta, tau_psi=tau_theta, 
                                   mu_psi=mu_psi, tau_psi=tau_psi))
    e = extract(s, c('theta','psi'))
    data.frame(theta = as.numeric(e$theta),
               psi = as.numeric(e$psi))
  })

  mu_theta = mean(draws$theta)
  tau_theta = mean((draws$theta-mu_theta)^2)
  mu_psi = mean(draws$psi)
  tau_psi = mean((draws$psi-mu_psi)^2)
  keep$mu_theta[i] = mu_theta; keep$tau_theta[i] = tau_theta
  keep$mu_psi[i] = mu_psi; keep$tau_psi[i] = tau_psi
}
@

<<dependson=c('normal_hierarchical_mean_em','normal_hierarchical_mean_mcem'), echo=FALSE>>=
ggplot(keep, id.var=c("iteration"), variable.name="parameter"), 
       aes(iteration, value, color=method)) +
  geom_point() + 
  facet_wrap(~parameter, scales="free_y") +
  geom_hline(data=melt(truth, variable.name='parameter'), aes(yintercept=value))
@
